---
title: "Generative model simulations"
output: html_document
---
# load packages and additional scripts
```{r}
source("function.R")
library(lfmm)
library(cate)
library(glmnet)
library(sva)
library(tidyverse)
library(reshape2)
```

#  Running generative simulations

This section reproduces the experiments described in section "Generative simulation experiments" of the manuscript.

## Loading the data and setting simulation parameters

Generative simulations are based on the generative model of lfmm.

```{r}
# Number of Individuals
n <- 400
# n <- 1000
# Number of Genetic markers
p <- 500 
# p <- 20000 # warning, take a long time to run
# Number of latent factors
k <- 6

# Number of causal genetic markers
nc <- 80 

# Hyperparameters
pc <- nc/p

# simulation parameters tested :
# proportion of exposure variance explained by latent structure (intensity of confounding)
prv <- rep(c(0.1, 0.5), 3)
# prv <- rep(seq(0.1, 0.7, 0.2), 200)
# mean of effect sizes (for causal markers)
eff <- c(0.75, 1.5)

para <- expand.grid(prv, eff)
colnames(para) <- c("prop_var", "eff")

# Number of simulation
nb <- nrow(para)

# Performances matrices
# F1-score
f1 <- matrix(NA, nrow = nb, ncol = 5)
# Root Mean Square Error computed on all markers
rmse <- matrix(NA, nrow = nb, ncol = 6)
# Root Mean Square Error computed on causal markers
rmse.c <- matrix(NA, nrow = nb, ncol = 5)
# Precision (1 - False Discovery Rate)
preci <- matrix(NA, nrow = nb, ncol = 5)
# Recall (Power)
recal <- matrix(NA, nrow = nb, ncol = 5)
```

## Launch all runs 

For each  simulation with the simulator_ewas() function, we perform GWAS with the following 6 methods: sparse lfmm, bslmm, lasso, ridge lfmm, cate and sva. Then for each method, we calculate the F1-score, the precision and the recall.


```{r}


for (j in 1:nrow(para)) {
  
  # latent factors with very small effect
  kr <- 4

    simu <- simulator_ewas(n = n, p = p, K = k + kr, prop.causal = pc, 
                           prop.variance = para$prop_var[j], mean.A = para$eff[j], 
                           sd.V = 1, sd.U = c(runif(k, 2, 6), runif(kr, 0, 1)))
    
    # for each method, we only conserve beta value
    
    # sparse lfmm
    slf <- lfmm::lfmm_lasso(Y = simu$Y, X = simu$X, K = k, nozero.prop = pc)$B

    # lasso from glmnet package
    cvfit = glmnet::cv.glmnet(x = simu$Y, y = simu$X)
    a <- coef(cvfit, s = cvfit$lambda.1se) 
    b <- data.frame(a@i, a@x)[-1,] # remove intercept
    las <- rep(0, p)
    las[b[,1]] <- b[,2]

    # lfmm ridge
    rlf <- lfmm::lfmm_ridge(Y = simu$Y, X = simu$X, K = k)$B

    # cate 
    cat <- cate::cate.fit(X.primary = simu$X, Y = simu$Y, r = k)$beta

    # sva
    mod <- model.matrix(~ simu$X)
    mod0 <- model.matrix(~ 1, data = data.frame(simu$X))
    svobj <- sva::sva(t(simu$Y), mod, mod0, n.sv = k)
    modSv <- cbind(mod, svobj$sv)
    # linear model between exposure and the matrix of genetic markers (with latent lariable estimate with sva)
    sva <- apply(simu$Y, 2, function(x) summary(lm(x ~ modSv[,-1]))$coefficients[2, 1])
    
    # set of causal markers
    causal <- simu$causal
  
  
    # F1 score calculation, precision and recall
    a.slf <- F1(slf, causal, nb.hit = 100)
    a.las <- F1(las, causal, nb.hit = 100)
    a.rlf <- F1(rlf, causal, nb.hit = 100)
    a.cat <- F1(cat, causal, nb.hit = 100)
    a.sva <- F1(sva, causal, nb.hit = 100)
    
    # RMSE calculation
    r.slf <- RMSE(slf, simu$A)
    r.las <- RMSE(las, simu$A)
    r.rlf <- RMSE(rlf, simu$A)
    r.cat <- RMSE(cat, simu$A)
    r.sva <- RMSE(sva, simu$A)
    r.zer <- RMSE(rep(0, p), simu$A)
    
    # RMSE calculation, only on causal markers
    rc.slf <- RMSE(slf[causal], simu$A[causal])
    rc.las <- RMSE(las[causal], simu$A[causal])
    rc.rlf <- RMSE(rlf[causal], simu$A[causal])
    rc.cat <- RMSE(cat[causal], simu$A[causal])
    rc.sva <- RMSE(sva[causal], simu$A[causal])

    # Result matrices
    # F1-score
    f1[j,] <- c(a.slf$f1, a.las$f1, a.rlf$f1, a.cat$f1, a.sva$f1)
    # Precision
    preci[j,] <- c(a.slf$precision, a.las$precision,a.rlf$precision, a.cat$precision, a.sva$precision)
    # Recall
    recal[j,] <- c(a.slf$recall, a.las$recall, a.rlf$recall, a.cat$recall, a.sva$recall)
    # Root Mean Square Error computed on all markers
    rmse[j,] <- c(r.slf, r.las, r.rlf, r.cat, r.sva, r.zer)
    # Root Mean Square Error computed on causal markers
    rmse.c[j,] <- c(rc.slf, rc.las, rc.rlf, rc.cat, rc.sva)
    
    print(paste0("=============="))
    print(paste0("============== ", j / nrow(rmse) * 100, " %"))
    print(paste0("=============="))

}

# Format the results
colnames(rmse) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva", "zero")
colnames(rmse.c) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(f1) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(preci) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(recal) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")

rmse <- data.frame(para, rmse)
rmse.c <- data.frame(para, rmse.c)
f1 <- data.frame(para, f1)
preci <- data.frame(para, preci)
recal <- data.frame(para, recal)

# res <- list(f1 = f1, rmse = rmse, preci = preci, recal = recal, rmse.c = rmse.c)
```

# Script for figure 1

Root Mean Square Error (RMSE) as a function of the effect size of causal markers and confounding intensity. Two sparse methods (sparse LFMM, LASSO) and three non-sparse methods (ridge LFMM, CATE and SVA) were compared. The “Zero” value corresponds to an RMSE obtained with all effect sizes set to zero (null-model error). Generative model simulation parameters: (A) Lower effect sizes and confounding intensities (B) Lower effect sizes and higher confounding intensities. (C) Higher effect sizes and lower confounding intensities. (D) Higher effect sizes and confounding intensities.

```{r}
rmse <- melt(rmse, id.vars = 1:2)
# proportion of variance parameter transformed into factor
# 0.1 --> Low Proportion of variance
# 0.5 --> High Proportion of variance
rmse$prop_var <- as.factor(rmse$prop_var)
levels(rmse$prop_var) <- c("LOW \n CONFOUNDING", "HIGH \n CONFOUNDING")
# Effect Size parameter transformed into factor
# 0.75 --> Low EFFECT SIZE
# 1.5 --> High EFFECT SIZE
rmse$eff <- as.factor(rmse$eff)
levels(rmse$eff) <- c("LOW \n EFFECT \n SIZE", "HIGH \n EFFECT \n SIZE")

levels(rmse$variable) <- c("Sparse \n LFMM", "LASSO", "Ridge \n LFMM", "CATE", "SVA", "Zero")

# Summarize the resuslt
dat_mean <- rmse %>%
  group_by(variable, eff, prop_var) %>%
  summarize(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

# plot
ggplot(dat_mean, aes(variable, mean, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.5, position = position_dodge(0.9)) +
  facet_grid(eff ~ prop_var, scales = "free_x") +
  xlab("INTENSITY OF CONFOUNDING") +
  ylab("RMSE") +
  labs(fill = "METHODS") +
  theme_bw()+
  theme(strip.text.x = element_text(size = 10, face = "bold", angle = 0),
        strip.text.y = element_text(size = 10, face = "bold", angle = 0),
        strip.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 15, face = "bold"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")+
  scale_fill_manual(values = c("orange", "purple", "blue", "green", "yellow", "black"))

```

# Script for figure 2

F-score and precision as a function of effect size of the causal markers and confounding intensity. Two sparse methods (sparse LFMM, LASSO) and three non-sparse methods (ridge LFMM, CATE and SVA) were com- pared. F-score is the harmonic mean of precision and recall. Generative model simulation parameters: (A) Lower effect sizes and confounding intensities (B) Lower effect sizes and higher confounding intensities. (C) Higher effect sizes and lower confounding intensities. (D) Higher effect sizes and confounding intensities.

```{r}
dat <- rbind(cbind(f1, stat = "F1 SCORE"),
             cbind(recal, stat = "RECALL"),
             cbind(preci, stat = "PRECISION"))

dat <- melt(dat, id.vars = c(1:2, 8))

# proportion of variance parameter transformed into factor
# 0.1 --> Low Proportion of variance
# 0.5 --> High Proportion of variance
dat$prop_var <- as.factor(dat$prop_var)
levels(dat$prop_var) <- c("LOW \n CONFOUNDING", "HIGH \n CONFOUNDING")
# Effect Size parameter transformed into factor
# 0.75 --> Low EFFECT SIZE
# 1.5 --> High EFFECT SIZE
dat$eff <- as.factor(dat$eff)
levels(dat$eff) <- c("LOW \n EFFECT \n SIZE", "HIGH \n EFFECT \n SIZE")

levels(dat$variable) <- c("Sparse \n LFMM", "LASSO", "Ridge \n LFMM", "CATE", "SVA")

# Summarize the resuslt
dat_mean <- dat %>%
  group_by(variable, eff, prop_var, stat) %>%
  summarize(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

filter(dat_mean, stat != "RECALL") %>%
  ggplot(., aes(variable, mean, fill = stat)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.5, position = position_dodge(0.9)) +
  facet_grid(eff ~ prop_var, scales = "free_x") +
  xlab("INTENSITY OF CONFOUNDING") +
  ylab("STATISTICS") +
  labs(fill = NULL) +
  theme_bw()+
  theme(strip.text.x = element_text(size = 10, face = "bold", angle = 0),
        strip.text.y = element_text(size = 10, face = "bold", angle = 0),
        strip.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 15, face = "bold"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 14, angle = 0),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom") +
  #scale_fill_manual(values = c("orange", "yellow", "blue", "green", "purple", "black")) +
  scale_fill_manual(values = c("black", "grey")) +
  scale_y_continuous(limits = c(0, 1))

```

